Approximation as Differentiation: Notes

### 2025-05-26 main/category-of-families.tex
First pass over FinVect, still a bit subscript-heavy, but better.

Following the CHAD approach, automatic differentiation for first-order programs can also be recovered in this
setup, via the total category $\Fam(\FinVect)$. We recapitulate the basic idea here. For flat manifolds like
$\RR^n$, the tangent space and cotangent space at a point $x$ are both canonically isomorphic to $\RR^n$, so
the forward and backward derivatives of a differentiable function $f: \RR^m \to \RR^n$ at a point $x$ have the
following form:

>>>>

Following the CHAD approach, automatic differentiation for first-order programs can also be recovered in this
setup, via the total category $\Fam(\FinVect)$. We recapitulate the basic ideas here, starting with the
category $\FinVect$ of finite vector spaces over $\RR$, with addition of real numbers written $a + b$ and
multiplication written $a \mult b$.

\subsubsection{Category of finite-dimensional vector spaces}
\label{sec:categories-with-biproducts:fdvect}

\begin{definition}[Finite dimensional vector space over $\RR$]
Up to isomorphism, an \emph{$n$-dimensional vector space over $\RR$} is the set $\RR^n$ equipped with
component-wise addition of vectors $+: \RR^n \times \RR^n \to \RR^n$ and scalar multiplication $\mult: \RR
\times \RR^n \to \RR^n$ (overloading $+$ and $\mult$), where $(\RR^n,+)$ is an abelian group (with identity
$0$ and additive inverse $-v$ again defined component-wise) and the vector operations are compatible with the
field operations in that the following equations hold:

\vspace{-4mm}
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{align*}
   1 \mult v &= v \\
   (a \mult b) \mult v &= a \mult (b \mult v)
\end{align*}
\end{center}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{align*}
   (a + b) \mult v &= (a \mult v) + (b \mult v) \\
   a \mult (u + v) &= (a \mult u) + (b \mult v)
\end{align*}
\end{center}
\end{minipage}
\end{definition}

\begin{definition}[Category $\FinVect$]
Define $\FinVect$ to be the category which has as objects $\RR^n$ all finite-dimensional vector spaces, and as
morphisms $f: \RR^m \to \RR^n$ all functions $f$ satisfying $f(u + v) = f(u) + f(v)$ and $f(a \mult v) = a
\mult f(v)$.
\end{definition}

$\FinVect$ is enriched in $\Ab$, the category of abelian groups. For any objects $V = \RR^m$ and $W = \RR^n$
the hom-set $\FinVect(V,W)$ is an abelian group, with identity $0_{V,W}$ given by the constant map $v \mapsto
0_W$ and addition and additive inverse given pointwise:

\vspace{-4mm}
\begin{center}
\begin{minipage}[t]{0.45\textwidth}
\begin{align*}
(f + g)(v) &= f(v) + g(v)
\end{align*}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{align*}
(-f)(v) &= -f(v)
\end{align*}
\end{minipage}
\end{center}

\vspace{1mm}
\noindent Composition is with bilinear. The direct sum $V \biprod W \iso \RR^{m + n}$ given by the Cartesian
product is a biproduct $(V \biprod W, \biinj_{V}, \biinj_{W}, \pi_1, \pi_2)$ with $\biinj_{V} =
\prodM{\id_{V}}{0_{V,W}}$ and $\biinj_{W} = \prodM{0_{V,W}}{\id_{W}}$.

\subsubsection{Derivatives of Differentiable Functions}

For flat manifolds like $\RR^n$, the tangent space and cotangent space at a point $x$ are both canonically
isomorphic to $\RR^n$, so the forward and backward derivatives of a differentiable function $f: \RR^m \to
\RR^n$ at a point $x$ have the following form:
